---
title: "Chapter 10. Counting and Classification"
output: github_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)

library(glue)
library(broom)
library(patchwork)
library(rethinking)
library(tidyverse)
library(conflicted)


conflict_prefer("filter", "dplyr")
conflict_prefer("select", "dplyr")
conflict_prefer("rename", "dplyr")

# To be able to use `map2stan()`:
conflict_prefer("collapse", "dplyr")
conflict_prefer("extract", "tidyr")
conflict_prefer("rstudent", "rethinking")
conflict_prefer("lag", "dplyr")
conflict_prefer("map", "purrr")
conflict_prefer("Position", "ggplot2")

theme_set(theme_minimal())
source_scripts()
set.seed(0)
```

- information if often thrown away by using ratios of counts instead of the counts themselves
    * 10/20 and 1/2 are the same ratio, but the first has more information
- there is friction in using the count data instead of the proportions
    * interpretation is less intuitive
- this chapter covers the two most popular count distributions:
    1. *binomial regression*: model a binary classification
        - dead/alive, accept/reject
    2 *poisson regression*: models outcome without a known maximum
        - a binomial models with a very large maximum and samll probability per trial
        - number of elephants in Kenya, number of people who apply to graduate school

## 10.1 Binomial regression

- the following formula where $y$ is a count, $p$ is the probability of success in a trial, $n$ is the number of trials
$$
y \sim \text{Binomial}(n, p)
$$

- the two most common GLMs that use binomial likelihood functions:
    1. *logistic regression*: data are in single-trial cases and the outcome is only 0 or 1
    2. *aggregated binomial regression*: when individual trials with the same covariate values are aggregated together
        - the outcome can take a value of 0 or any positive integer up to $n$ number of trials
- both GLMs use the same logit link function
    * so both are sometimes called logistic regression and they can be converted between each other

## 10.1.1 Logisitc regression: Prosocial chimpanzees

- example experimental data
    * measure the prosocial behaviour of chimps
    * a focal chimp has the option to pull two levers where the first only gives the focal chimp food and the other gives the focal chimp food, but not the other chimp
    * therefore, the focal chimp always gets food and decides if the other chimp gets food
    * the control condition is to not have another chimp and the partner condition is to have the second chimp
    * the two choices are swapped from the left and right hand sides to detect any handedness of the focal chimps

```{r}
data("chimpanzees")
d <- as_tibble(chimpanzees) %>% janitor::clean_names()
d
```

- we will focus on the columns:
    * `pulled_left`: the outcome to predict
    * `prosoc_left`: a predictor for if the left-hand lever was the prosocial option
    * `condition`: contains a 1 for when there was a second partner chimp
- the model we will fit:
    * $L$: indicates if the left-hand lever was pulled
    * $P$: indicates if the left-hand option was pro-social
    * $C$: indicates whether or not the condition was with the partner

$$
L_i \sim \text{Binomial}(1, p_i) \\
\text{logit}(p_i) = \alpha + (\beta_P + \beta_{PC} C_i)P_i \\
\alpha \sim \text{Normal}(0, 10) \\
\beta_P \sim \text{Normal}(0, 10) \\
\beta_PC \sim \text{Normal}(0, 10)
$$

- this model includes an interaction term for the left-hand option being pro-social and whether or not there is a second chimp
    * also there is no main effect of the `condition` $C_i$ because we do not expect the precense of a second chimp on its own to predict whether the focal chimp pulls the left lever
- the priors are gently regularizing
- as comparitive measures of overfitting, fit two other models with fewer parameters
    * one with just an intercept
    * one without only th `prosoc_left` predictor (no predictor for `condition` whether there is a second chimp)

$$
L_i \sim \text{Binomial}(1, p_i) \\
\text{logit}(p_i) = \alpha \\
\alpha \sim \text{Normal}(0, 10)
$$

$$
L_i \sim \text{Binomial}(1, p_i) \\
\text{logit}(p_i) = \alpha + \beta_P P_i \\
\alpha \sim \text{Normal}(0, 10) \\
\beta_P \sim \text{Normal}(0, 10)
$$

- first we will inspect the simplest model, the one with only an intercept

```{r}
m10_1 <- quap(
    alist(
        pulled_left ~ dbinom(1, p),
        logit(p) <- a,
        a ~ dnorm(0, 10)
    ),
    data = d
)

precis(m10_1)
```

- $\alpha$ is on the scale of log-odds
    * to get it to probability scale, must use the inverse link function, the logistic
    * the 'rethinking' package offers the function `logisitic()` to do this, but below I just show the calculation for education reasons

```{r}
# MAP
1 / (1 + exp(-0.32))

# 89% interval
c(1 / (1 + exp(-0.18)), 1 / (1 + exp(-0.46)))
```

```{r}
logistic
```

- $\text{logistic}(0.32) \approx 0.58$ means that the probability of pulling the left-hand lever was 0.58 with an 89% interval of 0.54 to 0.61
    * the chimps had a tendency to favor the left without any other information
- the following two code chunks fit the other two models proposed above

```{r}
m10_2 <- quap(
    alist(
        pulled_left ~ dbinom(1, p),
        logit(p) <- a + bp*prosoc_left,
        a ~ dnorm(0, 10),
        bp ~ dnorm(0, 10)
    ),
    data = d
)

precis(m10_2)
```

```{r}
m10_3 <- quap(
    alist(
        pulled_left ~ dbinom(1, p),
        logit(p) <- a + (bp + bpc*condition)*prosoc_left,
        a ~ dnorm(0, 10),
        bp ~ dnorm(0, 10),
        bpc ~ dnorm(0, 10)
    ),
    data = d
)

precis(m10_3)
```

```{r}
compare(m10_1, m10_2, m10_3)
plot(compare(m10_1, m10_2, m10_3))
```

- from the WAIC, we can see that `m10_3` likely overfits a bit because its WAIC is greater than `m10_2`
    * though the difference in WAIC is samll, the difference standard error `dSE` is very small and suggests it is a real difference
- but `m10_3` should not just be rejected, it still reflects the structure of the experiment
    * we do want to see why `m10_3` performs worse than `m10_2`
- the estimates for `m10_3` show a negative interaction term with a large 89% interval
    * suggests the chimps don't care too much about the presence of another chimp
    * they do prefer to pull the prosocial option, though, because that estimate is 0.61 with an 89% interval well above 0
- to understand the impact of the estimate 0.61 for `bp`, must distinguish between the *absolute effect* and the *relative effect*
    * *absolute effect*: the change in the probability of the outcome, depending on all of the parameters
        - tells us the practical impact of a change in a predictor
    * *relative effect*: the proportional changes induced by a change in the predictor
        - the author claims that this effect can be misleading because they ignore the other parameters
- the relative effect:
    * consider the relative effect size of `prosoc_left` and its parameter `bp`
    * the customary measure of relativ effect for logisitic model is the *proportional change in odds*
        - just the exponent of the parameter estimate
        - it is $\exp(0.61) \approx 1.84$ for `bp`
        - odds are the ratio of the probability an even happens to the probability that it does not
    * for `bp`, the logodds of pulling the left-hand level (the outcome variable) is increased by 0.61
        - alternatively, the odds are multiplied by 1.84
    * the difficulty with proportional odds is that the actual change in probability depends on the intercept and the other predictor variables
        - for example, consider that the intercept $\alpha = 4$, then the probability of pulling the left-lever, ignoring all else, is $\text{logistic}(4) = 0.98$ 
        - then the increase from `bp` would be $\text{logistic}(4 + 0.61) = 0.99$
        - the difference from `bp` is really not very much on the absolute scale
- the absolute effect:
    * consider the model-average posterior predictive check to get a sense of the absolute effect of each treatment on the probability of pulling the left-hand lever
        - use th `ensemble()` function to take a weighted average (by WAIC) over the three models
        - in the following plot, this is compared to the proportion of times the left-hand lever was pull by each chimp in all four conditions
    * interpreting the plot
        - the chimps, on average, tended to pull the prosocial option on the left ("1,0" and "1,1")
        - the partner condition did not seem to matter because the heights of the lines did not tend to move when the condition was changed and hand of the lever was held constant
    

```{r}
d_pred <- tibble(
    prosoc_left = c(0, 1, 0, 1),
    condition = c(0, 0, 1, 1)
)

# Build an ensemble from all three models weighted by WAIC.
chimp_ensemble <- ensemble(m10_1, m10_2, m10_3, data = d_pred)

# Summarize the predictions.
pred_p <- apply(chimp_ensemble$link, 2, mean)
pred_p_pi <- apply(chimp_ensemble$link, 2, PI)

pred_tibble <- d_pred %>%
    mutate(pred_p = pred_p) %>%
    bind_cols(pi_to_df(pred_p_pi)) %>%
    mutate(group = paste(prosoc_left, condition, sep = ","),
           group = factor(group))
```


```{r}
chimp_data <- d %>%
    group_by(prosoc_left, condition, actor) %>%
    summarise(p = mean(pulled_left)) %>%
    ungroup() %>%
    mutate(group = paste(prosoc_left, condition, sep = ","),
           group = factor(group))

chimp_data %>%
    ggplot() +
    geom_line(aes(x = group, y = p, group = actor),
              color = "skyblue2") +
    geom_line(data = pred_tibble,
              aes(x = group,  y = pred_p, group = "1"),
              color = "black") +
    geom_ribbon(data = pred_tibble,
              aes(x = group, group = "1",
                  ymin = x5_percent, ymax = x94_percent),
              alpha = 0.2) +
    labs(x = "prosoc_left, condition",
         y = "proportion pulled left",
         title = "",
         subtitle = "Each blue line is a separate chimp in each combination of 'prosoc_left' and 'condition'.")
```

- the predictions are quite poor because they are averages across all chimps
    * a lot of variation among individuals could mask the association of interest
    * we can model this variation between individuals
- the chimps showed signs of handedness - some preferred pulling the left lever and others perferred pulling the right lever
    * estimate handedness as a distinct intercept for each chimp
- below is the model to fit
    * the intercept $\alpha$ has a subscript, one for each chimp
    * $alpha$ is a vector of parameters

$$
L_i \sim \text{Binomial}(1, p_i) \\
\text{logit}(p_i) = \alpha_{\text{ACTOR}[i]} + (\beta_P + \beta_{PC} C_i) P_i \\
\alpha_text{ACTOR} \sim \text{Normal}(0, 10) \\
\beta_P \sim \text{Normal}(0, 10) \\
\beta_PC \sim \text{Normal}(0, 10)
$$

- this model is coded and fit using MCMC below

```{r}
# Clean up data frame for use with `map2stan()`.
d2 <- d %>% select(pulled_left, actor, condition, prosoc_left)

m10_4 <- map2stan(
    alist(
        pulled_left ~ dbinom(1, p),
        logit(p) <- a[actor] + (bp + bpc*condition)*prosoc_left,
        a[actor] ~ dnorm(0, 10),
        bp ~ dnorm(0, 10),
        bpc ~ dnorm(0, 10)
    ),
    data = d2, chains = 2, iter = 2500, warmup = 500
)

precis(m10_4, depth = 2)
plot(m10_4)
```

- the posterior is not Guassian
    * e.g. the distribution for `a[2]`
        - the values are all positive, indicating a left-hand bias

```{r}
post <- extract.samples(m10_4)

tibble(a_2 = post$a[, 2]) %>%
    ggplot(aes(x = a_2)) +
    geom_density(fill = "grey50", alpha = 0.1) +
    labs(x = "alpha[actor == 2]", y = "probability density",
         title = "Posterior distribution for alpha when actor is 2")
```

- plotting the posterior predictive plots for each of the chimps shows how the intercept changes for each

```{r}
chimp <- 1

d_pred <- tibble(
    pulled_left = rep(0, 4), 
    prosoc_left = c(0, 1, 0, 1), 
    condition = c(0, 0, 1, 1)
)

d_pred_all <- tibble(
    actor = 1:7,
    data = rep(list(d_pred), 7)
) %>%
    unnest(data)

link_m10_4 <- link(m10_4, data = d_pred_all) %>% 
    as.data.frame() %>%
    as_tibble()
link_names <- c("0/0", "1/0", "0/1", "1/1")
link_names <- map(1:7, ~ paste0("chimp", .x, "_", link_names)) %>%
    unlist()
colnames(link_m10_4) <- link_names

link_m10_4 %>%
    mutate(sample_idx = row_number()) %>%
    pivot_longer(-sample_idx, names_to = "name", values_to = "value") %>%
    group_by(name) %>%
    summarise(avg = mean(value),
              pi = list(PI(value) %>% pi_to_df())) %>%
    ungroup() %>%
    unnest(pi) %>%
    mutate(actor = str_extract(name, "(?<=chimp)[:digit:]"),
           actor = as.numeric(actor),
           name = str_remove_all(name, "chimp[:digit:]_"),
           name = factor(name)) %>%
    ggplot() +
    facet_wrap(~ actor) +
    geom_line(aes(x = name, y = avg, group = "1"),
              size = 0.8, color = "black") +
    geom_ribbon(aes(x = as.numeric(name), 
                    ymin = x5_percent, ymax = x94_percent),
                alpha = 0.2, fill = "black") +
    scale_y_continuous(limits = c(0, 1.1)) +
    labs(x = "presoc_left/condition",
         y = "proportion pulled left",
         title = "A model with a different intercept for each chimp",
         subtitle = "Each subplot shows the proportion of times the left lever was pulled for a single chimp.")
```

### 10.1.2. Aggregated binomial: Chimpanzees again, condensed

- above, we looked at the proportion of times the chimp pulled the left-hand lever for each set of predictors
    * could also just count the number of pulls, as long as we don't care about the sequence

```{r}
d <- as_tibble(chimpanzees)
d_aggregated <- d %>%
    group_by(prosoc_left, condition, actor) %>%
    summarise(x = sum(pulled_left)) %>%
    ungroup()
d_aggregated
```

- can define the model using these counts
    * there were 18 trials for each animal

```{r}
m10_5 <- quap(
    alist(
        x ~ dbinom(18, p),
        logit(p) <- a + (bp + bpc*condition)*prosoc_left,
        a ~ dnorm(0, 10),
        bp ~ dnorm(0, 10),
        bpc ~ dnorm(0, 10)
    ),
    data = d_aggregated
)

precis(m10_5)
```

### 10.1.3 Aggregated binomial: Graduate school admissions

- often, the number of trials per condition is not constant
    * in these cases, must use another variable for the first parameter in `dbinom()`
- use UC Berkeley admission data as an example
    * only 12 rows with information about admissions to 6 departments, separated by male and female applicants

```{r}
data("UCBadmit")
d <- as_tibble(UCBadmit) %>% janitor::clean_names()
d
```

- goal: to estimate if there is gender bias in the admissions
- fit two models:
    * a binomial regression that models `admit` as a function of each applicant's gender
        - estimates the association between gender and probability of admission
    * a binomial regression that models `admit` as a constant, ignoring gender
        - this provides a sense of any overfitting in the first model
- below is the formula for the first model
    * $n_{\text{admit},i}$: the applications indexed by row number
    * $m_i$: a dummy variable for male (1) vs. female (0)

$$
n_{\text{admit},i} \sim \text{Binomial}(n_i, p_i) \\
\text{logit}(p_i) = \alpha + \beta_m m_i \\
\alpha \sim \text{Normal}(0, 10) \\
\beta_m \sim \text{Normal}(0, 10)
$$

```{r}
d$male <- as.numeric(d$applicant_gender == "male")

m10_6 <- quap(
    alist(
        admit ~ dbinom(applications, p),
        logit(p) <- a + bm*male,
        a ~ dnorm(0, 10),
        bm ~ dnorm(0, 10)
    ),
    data = d
)

m10_7 <- quap(
    alist(
        admit ~ dbinom(applications, p),
        logit(p) <- a,
        a ~ dnorm(0, 10)
    ),
    data = d
)

precis(m10_6)
precis(m10_7)

compare(m10_6, m10_7)
plot(compare(m10_6, m10_7))
```

- interpretation of the models:
    * the WAIC indicates that including the gender created a better model
    * this indicates that the gender matters a lot
        - being a male is an advantage: $exp(0.61) \approx 1.84$
        - the male applicant's odds were 184% that of a female's
    * the difference on the absolute scale is shown below

```{r}
post <- extract.samples(m10_6)
p_admit_male <- logistic(post$a + post$bm)
p_admit_female <- logistic(post$a)

diff_admit <- p_admit_male - p_admit_female
quantile(diff_admit, c(0.025, 0.50, 0.975))
```

- plot posterior predictions for the model
    * can use the function `postcheck()`, though I also made a plot of the same data

```{r}
postcheck(m10_6, n = 1e4)
```


```{r}
pred <- link(m10_6)

pred_tib <- tibble(avg = apply(pred, 2, mean)) %>%
    bind_cols(apply(pred, 2, PI) %>% pi_to_df())

d %>%
    mutate(case = factor(row_number()),
           prop_admit = admit / applications) %>%
    bind_cols(pred_tib) %>%
    ggplot(aes(x = case)) +
    facet_wrap(~ dept, scales = "free_x", nrow = 1) +
    geom_line(aes(y = prop_admit, group = dept)) +
    geom_point(aes(y = prop_admit, color = applicant_gender)) +
    geom_linerange(aes(ymin = x5_percent, ymax = x94_percent), alpha = 0.5) +
    geom_point(aes(y = avg), shape = 1) +
    scale_color_brewer(palette = "Set2") +
    labs(x = "case", y = "proportion admitted",
         title = "Admission proportions per department",
         subtitle = "Admission data are separated by gender of applicant.")
```

- from this plot we can see that there were only 2 deptartments with ower admission for females, yet the model says females should have a 14% lower chance of admission
    * the problem is that the departments that take the most students had fewer applications from females
- change our question:
    * previous question: "What are the average probabilties of admission for females and males *across all departments*?"
    * new question: "What is the average *difference* in probability of admission between females and males *within departments*?"
        - fit each department with its own intercept

$$
n_{\text{admit},i} \sim \text{Binomial}(n_i, p_i) \\
\text{logit}(p_i) = \alpha_{\text{DEPT}[i]} + \beta_m m_i \\
\alpha_{\text{DEPT}} \sim \text{Normal}(0, 10) \\
\beta_m \sim \text{Normal}(0, 10)
$$
- this model and one without accounting for gender (to check for overfitting) are fit below

```{r}
d$dept_id <- as.numeric(factor(d$dept))

m10_8 <- quap(
    alist(
        admit ~ dbinom(applications, p),
        logit(p) <- a[dept_id],
        a[dept_id] ~ dnorm(0, 10)
    ),
    data = d
)

m10_9 <- quap(
    alist(
        admit ~ dbinom(applications, p),
        logit(p) <- a[dept_id] + bm*male,
        a[dept_id] ~ dnorm(0, 10),
        bm ~ dnorm(0, 10)
    ),
    data = d
)

precis(m10_8, depth = 2)
precis(m10_9, depth = 2)
compare(m10_6, m10_7, m10_8, m10_9)
plot(compare(m10_6, m10_7, m10_8, m10_9))
```

- now the best model is the one with a different intercept for each department and no male predictor
    * still, the one with the different intercepts and male has some of the weight
    * now the odds are in favor of female admission with males have about 90% the odds of admission as a female in the same department

```{r}
exp(m10_9@coef[["bm"]])
```

```{r}
postcheck(m10_9)
```

```{r}
pred <- link(m10_9)

pred_tib <- tibble(avg = apply(pred, 2, mean)) %>%
    bind_cols(apply(pred, 2, PI) %>% pi_to_df())

d %>%
    mutate(case = factor(row_number()),
           prop_admit = admit / applications) %>%
    bind_cols(pred_tib) %>%
    ggplot(aes(x = case)) +
    facet_wrap(~ dept, scales = "free_x", nrow = 1) +
    geom_line(aes(y = prop_admit, group = dept)) +
    geom_point(aes(y = prop_admit, color = applicant_gender)) +
    geom_linerange(aes(ymin = x5_percent, ymax = x94_percent), alpha = 0.5) +
    geom_point(aes(y = avg), shape = 1) +
    scale_color_brewer(palette = "Set2") +
    labs(x = "case", y = "proportion admitted",
         title = "Admission proportions per department",
         subtitle = "Admission data are separated by gender of applicant.")
```

### 10.1.4 Fitting binomial regressions with `glm()`

```{r}
m10_9glm <- glm(cbind(admit, reject) ~ male + dept, 
                data = d, 
                family = binomial)

summary(m10_9glm)
```

## 10.2 Poisson regression

- when a binomial has a small probability of an event $p$ and a large number of trials $n$
    * a binomial has an expected value of $np$ and a variance $np(1-p)$
    * when $n$ is large and $p$ small, these become about the same
- example:
    * employ 1000 monks in a monastery to copy manuscripts (before the printing press)
    * on average, one monk finishes a manuscript per day
    * but each monk is working independently and the manuscripts vary in length
    * so some days 3 manuscripts finish, but many days it is none
    * variance: $np(1-p) = 1000(0.001)(1-0.001) \approx 1$
    * this is simulated with $1 \time 10^6$ monks below

```{r}
set.seed(0)
y <- rbinom(1e6, 1000, 1/1000)
c(mean(y), var(y))
```

- this special binomial is a *Poisson distribution*
    * useful for modeling binomial events with an unknown or very large number of trials $n$
- the model form for a Poisson is even simpler than for a binomial or Gaussian
    * because there is only one parameter

$$
y \sim \text{Poisson}(\lambda_i)
$$

- for the GLM, the link function is the log link
    * the log link makes $\lambda_i$ always positive
    * also implies an exponential relationship between predictors and the expected value

$$
y \sim \text{Poisson}(\lambda_i) \\
\log(\lambda_i) = \alpha + \beta x_i
$$

- $\lambda_i$ is the expected value and commonly thought of as a *rate*
    * allows to make models for the *exposure* varies across cases
- example:
    * one monastary is counting books completed per week, and another is counting per day
    * can analyze both in the same models even though the counts are aggregated over different amounts of time
    * treat $\lambda_i$ as the number of events $\mu$ per unit time (or distance) $\tau$; $\lambda_i = \mu / \tau$
    * the $\tau_i$ values are the different exposures
        - when $\tau_i = 1$ (of some unit) then $\log \tau_i = 0$ and the formula is identical to the first
        - when there are different values for the exposure (i.e. per day vs per week), this value currects for that

$$
y_i \sim \text{Poisson}(\lambda_i) \\
\log \lambda_i = \log(\frac{\mu_i}{\tau_i}) = \alpha + \beta x_i \\
\log \lambda_i = \log \mu_i - \log \tau_i = \alpha + \beta x_i \\
\log \mu_i = \log \tau_i + \alpha + \beta x_i \\
$$

### 10.2.1 Example: Oceanic tool complexity

- setup:
    * the old island societies of Oceania provide an example of technological evolution
        - they made fish hooks, axes, boats, hand plows, etc.
    * theorize that larger populations develop and sustain more complex tool kits
    * contact rates among populations increases population size, too
- the data and models:
    * `total_tools` is the outcome predictor
    * model the number of tools as the log of the `population`
    * the number of tools increases with `contact` rate
    * the impact of `population` counts is increased by high `contact` using a interaction term

```{r}
data("Kline")
d <- as_tibble(Kline) %>% 
    janitor::clean_names() %>%
    mutate(log_pop = log(population),
           contact_high = as.numeric(contact == "high"))
d
```

- the model formula:
    * $P$ is `population`, $C$ is `contact_high`
    * the priors are strongly regularizing due to the small amount of data

$$
T_i \sim \text{Poisson}(\lambda_i) \\
\log \lambda_i = \alpha + \beta+P \log P_i + \beta_C C_i + \beta_{PC} C_i \log P_i \\
\alpha \sim \text{Normal}(0, 100) \\
\beta_P \sim \text{Normal}(0, 1) \\
\beta_C \sim \text{Normal}(0, 1) \\
\beta_{PC} \sim \text{Normal}(0, 1)
$$

- fit the model with the quadratic approximation

```{r}
m10_10 <- quap(
    alist(
        total_tools ~ dpois(lambda),
        log(lambda) <- a + bp*log_pop + bc*contact_high + bpc*contact_high*log_pop,
        a ~ dnorm(0, 100),
        c(bp, bc, bpc) ~ dnorm(0, 1)
    ),
    data = d
)

precis(m10_10, corr = TRUE)
plot(precis(m10_10))
```

- interpretation:
    * the main effect of log-population `bp` is positive and both `bc` and `bpc` overlap zero substantially
    * could think that log-population is reliably associated with total tools, but that would be incorrect
        - easy to be mislead by tables of estimates, especially with interaction terms
- analyze the model by plotting counterfactual predictions
    * consider two islands with log-population of 8, but one is high contact and the other is low contact
    * calculate $\lambda$, the expected tool count, for each
        - sample from the posterior, convert with the linear model, take the exponential to reverse the logarithm
    * can plot the difference in the number of tools and find the percent of samples where the high contact group had more tools than the low contact group

```{r}
post <- extract.samples(m10_10)
lambda_high <- exp(post$a + post$bc + (post$bp + post$bpc)*8)
lambda_low <- exp(post$a + post$bp*8)

tibble(diff_vals = lambda_high - lambda_low) %>%
    ggplot(aes(x = diff_vals)) +
    geom_density(color = "skyblue4") +
    geom_vline(xintercept = 0, lty = 2, size = 0.8, color = "grey50") +
    labs(x = "lambda_high - lambda_low",
         y = "density")
```

```{r}
sum(lambda_high - lambda_low > 0) / length(lambda_high)
```

- there is a 95% plausibilty that the high-contact island has more tools than the low-contact island, holding population constant
    * suggests that contact is important even though the model estimates are not informative
    * this is because the uncertainty is `bc` and `bpc` are negatively correlated
        - when one is high, the other is low

```{r}
as_tibble(post) %>%
    ggplot(aes(x = bc, y = bpc)) +
    geom_point(alpha = 0.5, size = 0.3, color = "skyblue3")
```

- a better way to tell if a predictor is expected to improve prediction is to use model comparison

```{r}
# Without interaction
m10_11 <- quap(
    alist(
        total_tools ~ dpois(lambda),
        log(lambda) <- a + bp*log_pop + bc*contact_high,
        a ~ dnorm(0, 100),
        c(bp, bc) ~ dnorm(0, 1)
    ),
    data = d
)
# With only log-population.
m10_12 <- quap(
    alist(
        total_tools ~ dpois(lambda),
        log(lambda) <- a + bp*log_pop,
        a ~ dnorm(0, 100),
        bp ~ dnorm(0, 1)
    ),
    data = d
)
# With only contact rate.
m10_13 <- quap(
    alist(
        total_tools ~ dpois(lambda),
        log(lambda) <- a + bc*contact_high,
        a ~ dnorm(0, 100),
        bc ~ dnorm(0, 1)
    ),
    data = d
)
# Intercept only.
m10_14 <- quap(
    alist(
        total_tools ~ dpois(lambda),
        log(lambda) <- a,
        a ~ dnorm(0, 100)
    ),
    data = d
)
(islands_compare <- compare(m10_10, m10_11, m10_12, m10_13, m10_14))
plot(islands_compare)
```

- interpretation:
    * the top models include both predictors, but the weight is split 50:50 between the model without the interaction and the one with the interaction
        - indicates both predictors are informative
        - suggests the interaction is probably not important
- plot counterfactual predictions using the ensemble of the top 3 models
    * the input data is the log-population over the natural range with `contact_high` set to both 0 and 1
    * both trends curve up as log-pop increases
    * the impact of contact rate can be seen by the distance between the curves a little overlap of the 89% intervals

```{r}
log_pop_seq <- seq(6, 13, length.out = 30)
d_pred <- bind_rows(
    tibble(log_pop = log_pop_seq,
           contact_high = 1),
    tibble(log_pop = log_pop_seq,
           contact_high = 0)
)
lambda_pred <- ensemble(m10_10, m10_11, m10_12, data = d_pred)
lambda_med <- apply(lambda_pred$link, 2, median)
lambda_pi <- apply(lambda_pred$link, 2, PI) %>% pi_to_df()

d_pred %>%
    mutate(lambda_med,
           contact_high = factor(contact_high)) %>%
    bind_cols(lambda_pi) %>%
    ggplot(aes(x = log_pop)) +
    geom_ribbon(aes(ymin = x5_percent, ymax = x94_percent, 
                    fill = contact_high), 
                alpha = 0.2) +
    geom_line(aes(y = lambda_med, group = contact_high,
                  color = contact_high)) +
    geom_point(data = d, aes(y = total_tools, shape = factor(contact_high))) +
    scale_fill_brewer(palette = "Set1") +
    scale_color_brewer(palette = "Set1") +
    labs(x = "log-population", y = "total tools",
         color = "contact high", shape = "contact high",
         title = "Ensemble posterior predictions for the islands model set")
```

### 10.2.2 MCM islands

- verify that the MAP estimates made above accurately describe the shape of the posterior by fitting with MCMC

```{r}
m10_10_stan <- map2stan(m10_10, iter=3e3, warmup = 1e3, chains = 4)
precis(m10_10_stan)
plot(m10_10_stan)
```

- the MAP estimates are the same as before, so the posterior is approximately Gaussian
- look at pairs plots of the posterior distributions from the MCMC
    * samples of the intercept and coefficient for contact rate are high correlated
    * MCMC can help overcome problems from correlated coefficients, though it is still best to try to avoid them
    * try centering log-population and re-fitting `m10_10_stan`
    * this removes the correlation between the intercept and coefficient for `contact_high`
        - the chains were also more efficient (more *effective samples*; larger `n_eff`)

```{r}
extract.samples(m10_10_stan) %>%
    as_tibble() %>%
    sample_n(100) %>%
    GGally::ggscatmat(alpha = 0.5) +
    labs(title = "MCMC posterior samples from `m10_10_stan`")

d$log_pop_c <- d$log_pop - mean(d$log_pop)

m10_10_stan_c <- map2stan(
    alist(
        total_tools ~ dpois(lambda),
        log(lambda) <- a + bp*log_pop_c + bc*contact_high + bcp*log_pop_c*contact_high,
        a ~ dnorm(0, 10),
        bp ~ dnorm(0, 1),
        bc ~ dnorm(0, 1),
        bcp ~ dnorm(0, 1)
    ),
    data = d,
    iter = 3e3,
    warmup = 1e3,
    chains = 4
)

precis(m10_10_stan_c)
plot(m10_10_stan_c)

extract.samples(m10_10_stan_c) %>%
    as_tibble() %>%
    sample_n(100) %>%
    GGally::ggscatmat(alpha = 0.5) +
    labs(title = "MCMC posterior samples from `m10_10_stan_c`")
```

### 10.2.3 Example: Exposure and the offset

- use an example where the exposure varies across observations
    * e.g. length of observation, area of sampling, intensity of sampling
    * Poisson assumes the rate of events is constant in time or space
    * *offset*: add the logarithm of exposure to the linear model as another term
- simulated data:
    * own a monastery where the monks copy books by hand
    * we know the rate at which manuscripts are completed each day
    * suppose the true rate is $\lambda=1.5$ manuscripts per day

```{r}
set.seed(0)
num_days <- 30
y <- rpois(num_days, 1.5)
table(y)
```

- are thinking of purchasing another monastery, but want to know the productivity of the new monastery
    * but only have data for the other monastery on a per week basis
    * suppose the daily rate of the new monastery is actually $\lambda=0.5$ manuscripts per day
        - thus, the real per week rate if $\lambda = 0.5 \times 7 = 3.5$
```{r}
set.seed(0)
num_weeks <- 4
y_new <- rpois(num_weeks, 0.5*7)
table(y_new)
```

- build a data frame with data from both monasteries

```{r}
y_all <- c(y, y_new)
exposure <- c(rep(1, length(y)), rep(7, length(y_new)))
monastery <- c(rep(0, length(y)), rep(1, length(y_new)))
d <- tibble(y = y_all, days = exposure, monastery)
d
```

- fit the model to estimate the rate of manuscript production at each monastery
    * include log of exposure as a variable in the linear model

```{r}
d$log_days <- log(d$days)
m10_15 <- quap(
    alist(
        y ~ dpois(lambda),
        log(lambda) ~ log_days + a + b*monastery,
        a ~ dnorm(0, 100),
        b ~ dnorm(0, 1)
    ),
    data = d
)

precis(m10_15)
```

- compute the posterior distribution of $\lambda$ in each monastery
    * sample from the posterior and use the linear model *without* the offset
    * don't use the offset again because the parameters are already on the daily scale

```{r}
post <- extract.samples(m10_15)
lambda_old <- exp(post$a + post$b * 0)
lambda_new <- exp(post$a + post$b * 1)
precis(tibble(lambda_old, lambda_new))
```

```{r}
tibble(`old monastery` = lambda_old,
       `new monastery` = lambda_new) %>%
    pivot_longer(c("old monastery", "new monastery")) %>%
    ggplot(aes(x = value, color = name, fill = name)) +
    geom_density(alpha = 0.2) +
    scale_color_brewer(palette = "Set2") +
    scale_fill_brewer(palette = "Set2") +
    scale_y_continuous(expand = expansion(mult = c(0, 0.02))) +
    theme(legend.title = element_blank(),
          legend.position = c(0.8, 0.8)) +
    labs(x = "posterior samples of manuscripts per day (lambda)",
         y = "probability density",
         title = "Model the production rates of two monasteries")
```

- interpretation
    * the MAP of the two posteriors are pretty close to the known rates of manuscript production per day
    * the new monastery is predicted to produce 0.6 manuscripts per day while the old monastery produces 1.6 per day
    * the 89% intervals are well above 0 and not overlapping

## 10.3 Other count regressions

- binomial works when there are only 2 outcomes, but there are situations where we are counting more than two outcomes
- *multinomial distribution*: when more than two types of unordered events are possible and the probability of each type of event is constant across trials
    * the binomial is actually just a special case of this distribution
    * also called *categorical regression* or *maximum extropy classifier* (in ML)

$$
\Pr(y_1, ..., y_k | n, p_1, ..., p_K) = \frac{n!}{\Pi_i y_i !} \Pi_{i=1}^{K} p_i^{y_i}
$$

- two approaches to modeling the multinomial:
    1. *explicit approach*: directly uses the multinomial likelihood, and uses a generalization of the logit link function
    2. transform the multinomial likelihood into a series of Poisson likelihoods

#### 10.3.1.1 Explicit multinomial models

- use the multinomial logit as the link function
    * takes a vector of "scores", one for each of $K$ event types
    * computes the probability of a particular type of event $k$ as
    * available using the `softmax()` function from the 'rethinking' package

$$
\Pr(k | s_1, ..., s_K) = \frac{exp(s_k)}{\sum_{i=1}^{K} exp(s_i)}
$$

- use the multinomial logit to create a *multinomial logistic regression*
    * actually build $K-1$ linear models
    * each can use different predictors
- two types of cases when building a *multinomial logistic regression*:
    1. the predictors have different value for different types of events
        * useful when each type of event has its own traits
        * want to estimate the association of those traits wtih the probability of each type of event
    2. parameters are distinct for each type of event
        * useful when interesting in features of some entity that produces each event
- example of case 1: modeling choice of career for many people
    * a relevant predictor is expected income
        - $\beta_{\text{INCOME}}$ will be in each model
        - a different income value multiplies the parameter in each linear model
    * simulated data of 500 people with 3 career choices
        - each career has a different expected income

```{r}
N <- 500               # number of individuals
income <- 1:3          # expected income for each career
score <- 0.5 * income  # scores for each career based on income

# convert scores to probabilities
p <- softmax(score[1], score[2], score[3])

# Sample chosen careers for each person
set.seed(0)
career <- map_dbl(1:N, ~ sample(1:3, size = 1, prob = p))

tibble(career) %>%
    ggplot(aes(x = factor(career))) +
    geom_bar(fill = "skyblue3") +
    labs(x = "career", y = "count", 
         title = "Career choices of 500 individuals",
         subtitle = "Simulated data for multinomial logistic regression")
```

- fit the model with `dcategorical()` likelihood
    * the multinomial logistic regression distribution
    * works when each value of the outcome variable (`career`) contains individual event types on each row
- convert all the scores to probabilities using `softmax()`
    * this is the multinomial logit link
- one of the event types will be used as the reference type
    * this is why there are $K-1$ linear models (where $K$ is the number of event types)
    * instead of getting a linear model, this event type is assigned a constant
    * the other linear models contain parameters relative to the reference type

```{r}
m10_16 <- quap(
    alist(
        career ~ dcategorical(softmax(0, s2, s3)),
        s2 <- b*2,
        s3 <- b*3,
        b ~ dnorm(0, 5)
    ),
    data = list(career = career)
)

precis(m10_16)
```

- parameter estimates are difficult to interpret
    * must instead convert them to a vector of probabilities to make sense of them
    * the estimate's value depends upon which event type is assigned as the reference

```{r}
b_income <- m10_16@coef["b"]
tibble(
    career = 1:3,
    real = p,
    estimate = softmax(1*0, 2*b_income, 3*b_income)
)
```

- example of case 2:
    * still modeling career choice
    * now want to estimate the association between each person's family income and the career choice
    * the predictor variable will have the same value in each linear model, for each row of data
    * instead, there is a unique parameter multiplying it in each model
        - is the estimate of the impact of family income on the choice of career

```{r}
set.seed(0)

N <- 100                   # number of people
family_income <- runif(N)  # family income for each person
b <- seq(1, -1)            # real coefficient values

# Sample the career for each person
career <- map_dbl(1:N, function(i) {
    score <- 0.5 * 1:3 + b*family_income[i]
    p <- softmax(score[1], score[2], score[3])
    return(sample(1:3, size = 1, prob = p))
})

tibble(career) %>%
    ggplot(aes(x = factor(career), y = family_income)) +
    geom_boxplot(color = "grey50", fill = "skyblue3",
                 alpha = 0.3, width = 0.2) +
    ggbeeswarm::geom_quasirandom() +
    labs(x = "career", y = "family income", 
         title = "Career choices of 500 individuals",
         subtitle = "Simulated data for multinomial logistic regression")
```

```{r}
career_data <- tibble(career, family_income)

m10_17 <- quap(
    alist(
        career ~ dcategorical(softmax(0, s2, s3)),
        s2 <- a2 + b2*family_income,
        s3 <- a3 + b3*family_income,
        c(a2, a3, b2, b3) ~ dnorm(0, 5)
    ),
    data = career_data
)

precis(m10_17)
```

- again, the estimates are difficult to interpret without computing implied predictions

```{r}
pred_data <- tibble(
    career = rep(1:3, each = 10),
    family_income = rep(seq(0, 1, length.out = 10), 3)
)

pred <- sim(m10_17, pred_data)

pred_data %>%
    mutate(estimate = apply(pred, 2, mean)) %>%
    group_by(family_income) %>%
    mutate(estimate = softmax(estimate)) %>%
    ungroup() %>%
    arrange(family_income, career) %>%
    ggplot(aes(x = factor(career), y = family_income)) +
    geom_point(aes(size = estimate, color = estimate)) +
    scale_color_viridis_c() +
    labs(x = "career",
         y = "family income",
         title = "Probability of career choice given family income",
         subtitle = "Probabilities were estimated using a multinomial logisitic regression.",
         color = "estimate", size = "estimate")
```

#### 10.3.1.2 Multinomial in disguise as Poisson

