---
title: "Chapter 11. Monsters and Mixtures"
output: github_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE, cache = TRUE)

library(glue)
library(broom)
library(patchwork)
library(MASS)
library(rethinking)
library(tidyverse)
library(conflicted)


conflict_prefer("filter", "dplyr")
conflict_prefer("select", "dplyr")
conflict_prefer("rename", "dplyr")

# To be able to use `map2stan()`:
conflict_prefer("collapse", "dplyr")
conflict_prefer("extract", "tidyr")
conflict_prefer("rstudent", "rethinking")
conflict_prefer("lag", "dplyr")
conflict_prefer("map", "purrr")
conflict_prefer("Position", "ggplot2")

# To be able to use 'MASS':
conflict_prefer("area", "patchwork")

theme_set(theme_minimal())
source_scripts()
set.seed(0)
```

- build more types of models by piecing together types we have already learned about
    * will discuss *ordered categorical* models and *zero-inflated*/*zero-augmented* models
- mixtures are powerful, but interpretation is difficult

## 11.1 Ordered categorical outcomes

- the outcome is discrete and has different *levels* along a dimension but the differences between each level are not necessarily equal
    * is a multinomial prediction problem with a constraint on the order of the categories
    * want an estimate of the effect of a change in a predictor on the change along the categories
- use a *cumulative link function* 
    * the cumulative probability of a value is *the probability of that value or any smaller value*
    * this guarantees the ordering of the outcomes

### 11.1.1 Example: Moral intuition

- example data come from a survey of people with different versions of the classic "Trolley problem"
    * 3 versions that invoke different moral principles: "action principle," "intention principle," and "contact principle"
    * the goal is how people just the different choices from the different principles
    * `response`: from an integer 1-7, how morally permissible is the action

```{r}
data("Trolley")
d <- as_tibble(Trolley)
```

### 11.1.2 Describing an ordered distribution with intercepts

- some plots of the data
    * a histogram of the response values
    * cumulative proportion of responses
    * log-cumulative-odds of responses

```{r}
p1 <- d %>%
    ggplot(aes(x = response)) +
    geom_histogram(bins = 30) +
    labs("Distribution of response")

p2 <- d %>% 
    count(response) %>%
    mutate(prop = n / sum(n),
           cum_prop = cumsum(prop)) %>%
    ggplot(aes(x = response, y = cum_prop)) +
    geom_line() +
    geom_point() +
    labs(y = "cumulative proportion")

p3 <- d %>% 
    count(response) %>%
    mutate(prop = n / sum(n),
           cum_prop = cumsum(prop),
           cum_odds = cum_prop / (1 - cum_prop),
           log_cum_odds= log(cum_odds)) %>%
    filter(is.finite(log_cum_odds)) %>%
    ggplot(aes(x = response, y = log_cum_odds)) +
    geom_line() +
    geom_point() +
    labs(y = "log-cumulative-odds")

p1 | p2 | p3
```

- why use the log-cumulative-odds of each response:
    * it is the cumulative analog of the logit link used previously
    * the logit is the log-odds; the cumulative logit is the log-cumulative-odds
    * constrains the probabilities to between 0 and 1
    * this link function takes care of converting the parameter estimates to the probability scale
- to use Bayes' theorom to compute the posterior distribution of these intercepts, need to compute the likelihood of each possible response value
    * need to use the cumulative probabilities $\Pr(y_i \ge k)$ to compute the likelihood $\Pr(y_i = k)$
    * use the inverse link to translate the log-cumulative-odds back to cumulative probability
    * therefore, when we observe $k$ and need its likelihoood, just use subtraction:
        -  the values are shown as blue lines in the next plot

$$
p_k = \Pr(y_i = k) = \Pr(y_i \le k) - \Pr(y_i \le k - 1)
$$

```{r}
offset_subtraction <- function(x) {
    y <- x
    for (i in seq(1, length(x))) {
        if (i == 1) { 
            y[[i]] <- x[[i]]
        } else {
            y[[i]] <- x[[i]] - x[[i - 1]]
        }
    }
    return(y)
}

d %>% 
    count(response) %>%
    mutate(prop = n / sum(n),
           cum_prop = cumsum(prop),
           likelihood = offset_subtraction(cum_prop),
           ymin = cum_prop - likelihood) %>%
    ggplot(aes(x = response)) +
    geom_linerange(aes(ymin = 0, ymax = cum_prop), color = "grey50") +
    geom_line(aes(y = cum_prop)) +
    geom_point(aes(y = cum_prop)) +
    geom_linerange(aes(ymin = ymin, ymax = cum_prop), color = "blue",
                   position = position_nudge(x = 0.05), size = 1) +
    labs(y = "log-cumulative-odds",
         title = "Cumulative probability and ordered likelihood",
         subtitle = "Blue lines indicate the likelihood for each response.")
```

