---
title: "Chapter 11. Monsters and Mixtures"
output: github_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE, comment = "#>", cache = TRUE, dpi = 500)

library(glue)
library(mustashe)
library(broom)
library(patchwork)
library(MASS)
library(rethinking)
library(tidyverse)
library(conflicted)


conflict_prefer("filter", "dplyr")
conflict_prefer("select", "dplyr")
conflict_prefer("rename", "dplyr")

# To be able to use `map2stan()`:
conflict_prefer("collapse", "dplyr")
conflict_prefer("extract", "tidyr")
conflict_prefer("rstudent", "rethinking")
conflict_prefer("lag", "dplyr")
conflict_prefer("map", "purrr")
conflict_prefer("Position", "ggplot2")

# To be able to use 'MASS':
conflict_prefer("area", "patchwork")

theme_set(theme_minimal())
source_scripts()
set.seed(0)

# For 'rethinking'.
options(mc.cores = parallel::detectCores())
rstan_options(auto_write = TRUE)
```

- build more types of models by piecing together types we have already learned about
    * will discuss *ordered categorical* models and *zero-inflated*/*zero-augmented* models
- mixtures are powerful, but interpretation is difficult

## 11.1 Ordered categorical outcomes

- the outcome is discrete and has different *levels* along a dimension but the differences between each level are not necessarily equal
    * is a multinomial prediction problem with a constraint on the order of the categories
    * want an estimate of the effect of a change in a predictor on the change along the categories
- use a *cumulative link function* 
    * the cumulative probability of a value is *the probability of that value or any smaller value*
    * this guarantees the ordering of the outcomes

### 11.1.1 Example: Moral intuition

- example data come from a survey of people with different versions of the classic "Trolley problem"
    * 3 versions that invoke different moral principles: "action principle," "intention principle," and "contact principle"
    * the goal is how people just the different choices from the different principles
    * `response`: from an integer 1-7, how morally permissible is the action

```{r}
data("Trolley")
d <- as_tibble(Trolley) %>%
    rename(q_case = case)
```

### 11.1.2 Describing an ordered distribution with intercepts

- some plots of the data
    * a histogram of the response values
    * cumulative proportion of responses
    * log-cumulative-odds of responses

```{r}
p1 <- d %>%
    ggplot(aes(x = response)) +
    geom_histogram(bins = 30) +
    labs("Distribution of response")

p2 <- d %>% 
    count(response) %>%
    mutate(prop = n / sum(n),
           cum_prop = cumsum(prop)) %>%
    ggplot(aes(x = response, y = cum_prop)) +
    geom_line() +
    geom_point() +
    labs(y = "cumulative proportion")

p3 <- d %>% 
    count(response) %>%
    mutate(prop = n / sum(n),
           cum_prop = cumsum(prop),
           cum_odds = cum_prop / (1 - cum_prop),
           log_cum_odds= log(cum_odds)) %>%
    filter(is.finite(log_cum_odds)) %>%
    ggplot(aes(x = response, y = log_cum_odds)) +
    geom_line() +
    geom_point() +
    labs(y = "log-cumulative-odds")

p1 | p2 | p3
```

- why use the log-cumulative-odds of each response:
    * it is the cumulative analog of the logit link used previously
    * the logit is the log-odds; the cumulative logit is the log-cumulative-odds
    * constrains the probabilities to between 0 and 1
    * this link function takes care of converting the parameter estimates to the probability scale
- to use Bayes' theorom to compute the posterior distribution of these intercepts, need to compute the likelihood of each possible response value
    * need to use the cumulative probabilities $\Pr(y_i \ge k)$ to compute the likelihood $\Pr(y_i = k)$
    * use the inverse link to translate the log-cumulative-odds back to cumulative probability
    * therefore, when we observe $k$ and need its likelihoood, just use subtraction:
        -  the values are shown as blue lines in the next plot

$$
p_k = \Pr(y_i = k) = \Pr(y_i \le k) - \Pr(y_i \le k - 1)
$$

```{r}
offset_subtraction <- function(x) {
    y <- x
    for (i in seq(1, length(x))) {
        if (i == 1) { 
            y[[i]] <- x[[i]]
        } else {
            y[[i]] <- x[[i]] - x[[i - 1]]
        }
    }
    return(y)
}

d %>% 
    count(response) %>%
    mutate(prop = n / sum(n),
           cum_prop = cumsum(prop),
           likelihood = offset_subtraction(cum_prop),
           ymin = cum_prop - likelihood) %>%
    ggplot(aes(x = response)) +
    geom_linerange(aes(ymin = 0, ymax = cum_prop), color = "grey50") +
    geom_line(aes(y = cum_prop)) +
    geom_point(aes(y = cum_prop)) +
    geom_linerange(aes(ymin = ymin, ymax = cum_prop), color = "blue",
                   position = position_nudge(x = 0.05), size = 1) +
    labs(y = "log-cumulative-odds",
         title = "Cumulative probability and ordered likelihood",
         subtitle = "Blue lines indicate the likelihood for each response.")
```

- below is the matematical form of the model using an ordered logit likelihood
    * notation for these models can vary by author
    * the *Ordered* distribution is just a categorical distribution that takes a vector $\text{p} = \{p_1, p_2, p_3, p_4, p_5, p_6\}$
        - only 6 because the 7th level has the value 1 automatically
    * each response value $k$ gets an intercept parameter $\alpha_k$

$$
R_i \sim \text{Ordered}(p) \\
logit(P_k) = \alpha_k \\
\alpha_k \sim \text{Normal}(0, 10)
$$

- the first model does not include any predictor vairables
    * the link function is embedded in the likelihood function, already
        - simpler to type and makes the calculations more efficient, too
    * `phi` is a placeholder for now but will be used to add in predictor variables
    * the start values are included to start the intercepts in the right order
        - their exact values don't really matter, just the order

```{r}
m11_1 <- quap(
    alist(
        response ~ dordlogit(phi, c(a1, a2, a3, a4, a5, a6)),
        phi <- 0,
        c(a1, a2, a3, a4, a5, a6) ~ dnorm(0, 10)
    ),
    data = d,
    start = list(a1 = -2, a2 = -1, a3 = 0, a4 = 1, a5 = 2, a6 = 2.5)
)

precis(m11_1)
```

- transform from log-cumulative-odds to cumulative probabilities

```{r}
logistic(coef(m11_1))
```

```{r}
m11_1_link <- extract.samples(m11_1)
m11_1_link %>%
    as.data.frame() %>%
    as_tibble() %>%
    pivot_longer(tidyselect::everything()) %>%
    ggplot(aes(x = value, y = name, color = name, fill = name)) +
    ggridges::geom_density_ridges(size = 1, alpha = 0.4, ) +
    scale_color_brewer(palette = "Set1") +
    scale_fill_brewer(palette = "Set1") +
    theme(legend.position = "none") +
    labs(x = "posterior samples",
         y = "density",
         title = "Posterior probability distributions of intercept from categorical model")
```

### 11.1.3 Adding predictor variables

- to include predictor variables:
    * *define the log-cumulative-odds of each response $k$ as a sum of its intercept $\alpha_k$ and a typical linear model*
    * for example: add a predictor $x$ to the model
        - define the linear model $\phi_i = \beta x_i$
        - the cumulative logit becomes:

$$
\log \frac{\Pr(y_i \ge k)}{1 - \Pr(y_i \ge k)} = \alpha - \phi_i = \alpha - \beta x_i
$$

- this form keeps the correct ordering of the outcome values while still morphing the likelihood of each individual value as the predictor $x_i$ changes value
- the linear model $\phi$ is subtracted from the intercept:
    * because decreasing the log-cumulative-odds of every outcome value $k$ below the maximum shifts probability mass upwards towards higher outcome values
    * a positive $\beta$ value indicates than an increase in the predictor variable $x$ results in an increase in the average response
- for the Trolly data, we can icnlude predictor variables for the different types of questions: "action," "intention," and "contact"
    * the formulation of the log-cumulative-odds of each response $k$ is shown below
    * defines the log-odds of each possible response to be an additive model of the features of the story corresponding to each response

$$
\log \frac{\Pr(y_i \ge k)}{1 - \Pr(y_i \ge k)} = \alpha - \phi_i \\
\phi_i = \beta_A A_i + \beta_I I_i + \beta_C C_i
$$

```{r}
m11_2 <- quap(
    alist(
        response ~ dordlogit(phi, c(a1, a2, a3, a4, a5, a6)),
        phi <- bA*action + bI*intention + bC*contact,
        c(a1, a2, a3, a4, a5, a6) ~ dnorm(0, 10),
        c(bA, bI, bC) ~ dnorm(0, 10)
    ),
    data = d,
    start = list(a1 = -2, a2 = -1, a3 = 0, a4 = 1, a5 = 2, a6 = 2.5)
)
```

- fit another model with interactions between action and intention and between contact and intention
    * these two make sense in terms of the scenario we are modeling while an interaction between contact and action does not
        - contact is a type of action

```{r}
m11_3 <- quap(
    alist(
        response ~ dordlogit(phi, c(a1, a2, a3, a4, a5, a6)),
        phi <- bA*action + bI*intention + bC*contact + bAI*action*intention + bCI*contact*intention,
        c(a1, a2, a3, a4, a5, a6) ~ dnorm(0, 10),
        c(bA, bI, bC, bAI, bCI) ~ dnorm(0, 10)
    ),
    data = d,
    start = list(a1 = -2, a2 = -1, a3 = 0, a4 = 1, a5 = 2, a6 = 2.5)
)

coeftab(m11_1, m11_2, m11_3)
```

- interpretation:
    * the intercepts are difficult to interpret on their own, but act like regular intercepts in simpler models
        - they are the relative frequencies of the outcomes when all predictors are set to 0
    * there are 5 slope parameters: 3 main effects and 2 iinteractions
        - they are all far from 0 (can check with `precis(m)`)
        - they are all negative: each factor/interaction *reduces* the average response
        - these values are difficult to interpret as is, so they are investigated more below
- compare the models by WAIC
    * the model with interaction terms is sufficiently better than the other two, so we can safely proceed with just analyzing it and ignoring the other two

```{r}
compare(m11_1, m11_2, m11_3)
```

- plot implied predictions to understand what model `m11_3` implies
    * difficult to plot the predictions of log-cumulative-odds because each prediction is a vector of probabilities, one for each possible outcome
    * as a predictor variable changes value, the entire vector changes
- one common plot is to use the horiztonal axis for the predictor variable and the vertical axis for the cumulative probability
    * plot a curve for each response value

```{r}
post <- extract.samples(m11_3)

get_m11_3_predictions <- function(kA, kC, kI) {
    res <- tibble()
    for (s in 1:100) {
        p <- post[s, ]
        ak <- as.numeric(p[1:6])
        phi <- p$bA * kA + p$bI*kI + p$bC*kC + p$bAI*kA*kI + p$bCI*kC*kI
        pk <- pordlogit(1:6, a = ak, phi = phi)
        res <- bind_rows(
            res,
            tibble(lvl = 1:6, val = pk)
        )
    }
    return(res)
}


implied_predictions <- expand.grid(kA = 0:1, kC = 0:1, kI = 0:1) %>%
    as_tibble() %>%
    filter(!(kA == 1 & kC == 1)) %>% 
    group_by(kA, kC, kI) %>%
    mutate(preds = list(get_m11_3_predictions(kA, kC, kI))) %>%
    unnest(preds) %>%
    ungroup()

implied_predictions %>%
    mutate(facet = paste0("action: ", kA, ", ", "contact: ", kC)) %>%
    group_by(kA, kC, kI) %>%
    mutate(row_idx = row_number()) %>%
    ungroup() %>%
    mutate(line_group = paste(lvl, row_idx)) %>%
    ggplot(aes(x = kI, y = val, group = line_group)) +
    facet_wrap(~ facet, nrow = 1) +
    geom_hline(yintercept = 0:1, lty = 2, color = "grey70", size = 1) +
    geom_line(aes(color = factor(lvl)), alpha = 0.1) +
    scale_color_brewer(palette = "Dark2") +
    labs(x = "intention", 
         y = "probability",
         color = "level",
         title = "Implied predictions of model with interactions.")
```

- interpretation:
    * each horizontal line is the bounday between levels
    * the thickness of the boundary represents the variation in prediction
    * the change in height from intention changing from 0 to 1 indicates the predicted impact of changing a trolley story from non-intention to intention
        - the left-hand plot shows that there is not much change from switching intention when action and contact are both 0
        - the other two plots show the interaction between intention and the other variable that is set to 1
        - the middle plot shows that there is a large interaction between contact and intention

## 11.2 Zero-inflated outcomes

- *mixture model*: measurements arise from multiple proceesses; different causes for the same observation
    * uses more than one probability distribution
- common to need to use a mixture model for count variables
    * a count of 0 can often arise in more than one way
    * a 0 could occurr because nothing happens, because the rate of the event is very low, or because the event-generating process never began
    * said to be *zero-inflated*

### 11.2.1 Example: Zero-inflated Poisson

- previously, we use the monastery example to study the Poisson distribution
    * now imagine that the monks take breaks on some days and no manuscripts are made
    * want to estimate how often breaks are taken
- mixture model:
    * a zero can arise from two processes:
        1. the monks took a break
        2. the monks workked but did not complete a manuscript
    * let $p$ be the probability the monks took a break on a day
    * let $\lambda$ be the mean number of manuscripts completed when the monks work
- need a likelihood function that mixes these two processes:
    * the following equation says this: "The probability of observing a zero is the probability that the monks took a break OR ($+$) the probability the monks worked AND ($\times$) failed to finish."

$$
\Pr(0 | p, \lambda) = \Pr(\text{break} | p) + \Pr(\text{work} | p) \times \Pr(0 | \lambda) \\
\Pr(0 | p, \lambda) = p + (1-p) e^{-\lambda}
$$

- the likelihood of a non-zero value $y$ is below
    * the the probability that the monks work multiplied by the probability that the working monks produce a manuscript

$$
\Pr(y | p, \lambda) = \Pr(\text{break} | p)(0) + \Pr(\text{work} | p) \Pr(y | \lambda) \\
\Pr(y | p, \lambda) = (1 - p) \frac{\lambda^y e^{-\lambda}}{y!}
$$

- can define `ZIPoisson` as the distribution above with parameters $p$ (probability of 0) and $\lambda$ (mean of Poisson) to describe the shape
    * two linear models and two link functions, one for each process

$$
y_i \sim \text{ZIPoisson}(p_i, \lambda_i) \\
\text{logit}(p_i) = \alpha_p + \beta_p x_i \\
\log(\lambda_i) = \alpha_\lambda + \beta_\lambda x_i
$$

- need to simulate data for the monks taking breaks

```{r}
# They take a break on 20% of the days
prob_break <- 0.2 

# Average of 1 manuscript per working day.
rate_work <- 1

# Sample one year of production.
N <- 365

# Simulate which days the monks take breaks.
break_days <- rbinom(N, 1, prob = prob_break)

# Simulate the manuscripts completed.
y <- (1-break_days) * rpois(N, rate_work)
```

```{r}
tibble(y, break_days) %>%
    count(y, break_days) %>%
    mutate(break_days = ifelse(break_days == 0, "work", "break"),
           break_days = factor(break_days, levels = c("break", "work"))) %>%
    ggplot(aes(x = y, y = n)) +
    geom_col(aes(fill = break_days)) +
    scale_fill_manual(
        values = c("skyblue3", "grey50"),
        guide = guide_legend(title.position = "left",
                             title.hjust = 0.5,
                             label.position = "top",
                             ncol = 2)
    ) +
    theme(
        legend.position = c(0.7, 0.7)
    ) +
    labs(x = "number of manuscripts completed on a day",
         y = "count",
         fill = "Did the monks take a break?",
         title = "The number of manuscripts completed per day when monks can take breaks")
```

- now we can fit a model

```{r}
m11_4 <- quap(
    alist(
        y ~ dzipois(p, lambda),
        logit(p) <- ap,
        log(lambda) <- al,
        ap ~ dnorm(0, 1),
        al ~ dnorm(0, 10)
    ),
    data = tibble(y)
)

precis(m11_4)

logistic(m11_4@coef[["ap"]])
exp(m11_4@coef[["al"]])
```

- can get a very accurate prediction for the proportion of days taken off by the monks and the rate of manuscript production per working day
    * though, cannot determine whether the monks took any particular day off

## 11.3 Over-dispersed outcomes

- *over-dispersion*: the variance of a variable exceededs the expected amount for a model
    * e.g.: for a binomial, the expected value is $np$ and its variance $np(1-p)$
    * for a count model, this suggests that a necessary variable has been omitted
- ideally, would just include the missing variable to remove over-dispersion, but not always the case/possible
- two strategies for mitigating the effects of over-dispersion
    * use a *continuous mixture* model: a linear model is attached to a distribution of observations
        - common models: *beta-binomial* and *gamma-Poisson* (negative-binomial)
        - these are demonstrated in the following sections
    * employ a multilevel model (GLMM) and estimate the residuals of each observation and the distribution of those residuals
        - easier to fit than beta-binmial and gamma-Poisson GLMs
        - more flexible
        - handle over-dispersion and other kinds of heterogineity simulatneously
        - *GLMMs are covered in the next chapter*

### 11.3.1 Beta-binomial

- beta-binomial models assumes that each binomial count observation has its own probability of a success
    * estimates the distribution of probabilities of success across cases
        - instead of a single probability of success
    * predictor variables change the shape of this distribution instead of directly determining the probability of each success
- example: UCB admissions data
    * if we ignore the department, the data is very over-dispersed
        - because the departments vary a lot in baseline admission rates
    * therefore, ignoring the inter-department variation results in an incorrect inference about applicant gender
    * can fit a beta-binomial model, ignoring department
- a beta-binomial model will assume that each row of the data has a unique, unobserved probability of admission
    * these probabilities of admission have a common distribution described by the *beta distribution*
        - use the beta distribution because it can be used to calculate the likelihood function that averages over the unknown probabilities for each observation
    * beta-distribution has 2 parameters:
        1. $\bar{\textbf{p}}$: average probability
        2. $\theta$: shape parameter
    * shape parameter describes the spread of the distribution

```{r}
pbar <- 0.5
thetas <- c(1, 2, 5, 10, 20)
x_vals <- seq(0, 1, length.out = 50)
beta_dist_res <- tibble()
for (theta in thetas) {
    beta_dist_res <- bind_rows(
        beta_dist_res,
        tibble(theta = theta, 
               pbar = pbar, 
               x = x_vals,
               d = dbeta2(x_vals, pbar, theta))
    )
}

beta_dist_res %>%
    mutate(params = paste0("pbar: ", pbar, ", theta: ", theta)) %>%
    ggplot(aes(x = x, y = d)) +
    geom_line(aes(group = params, color = params)) +
    scale_color_brewer(palette = "Dark2") +
    labs(x = "probability", y = "density", color = NULL,
         title = "Beta-binomial distributions")
```

- we will build a linear model to $\bar{\textbf{p}}$ so that changes in predictor variables change the central tendency of the distribution
    * $A$ is the number of admissions (`admit` column)
    * $n$ is the number of applications (`applications` column)
    * any predictor variables could be included in the linear model for $\bar{p}$

$$
A_i \sim \text{BetaBinomial}(n_i, \bar{p}_i, \theta) \\
\text{logit}(\bar{p}_i) = \alpha \\
\alpha \sim \text{Normal}(0, 10) \\
\theta \sim \text{HalfCauchy}(0, 1)
$$

```{r}
data("UCBadmit")
d <- as_tibble(UCBadmit) %>%
    janitor::clean_names()

stash("m11_5", depends_on = "d", {
    m11_5 <- map2stan(
        alist(
            admit ~ dbetabinom(applications, pbar, theta),
            logit(pbar) <- a,
            a ~ dnorm(0, 2),
            theta ~ dexp(1)
        ),
        data = d,
        constraints = list(theta = "lower=0"),
        start = list(theta = 3),
        iter = 4e3, 
        warmup = 1e3, 
        chains = 2, 
        cores = 1
    )
})

plot(m11_5)
pairs(m11_5)
precis(m11_5)
```

- interpretation:
    * `a` is on the log-odds scale and defines $\bar{\textbf{p}}$ of the beta distribution of probabilities for each row of the data
    * therefore, the average probability of admission across departments is about 0.4, but the percentile is quite wide

```{r}
post <- extract.samples(m11_5)
quantile(logistic(post$a), c(0.025, 0.5, 0.975))
```

- to see what the model says of the data, need to account for correlation between $\bar{\textbf{p}}$ and $\theta$
    * these parameters define a distribution of distributions
    * 100 combinations of $\bar{\textbf{p}}$ and $\theta$ are shown in the following plot

```{r}
dist_of_dist <- map(1:300, function(i) {
    p <- logistic(post$a[i])
    theta <- post$theta[i]
    x <- seq(0, 1, length.out = 100)
    probs <- dbeta2(x, p, theta)
    return(tibble(i, p, theta, x, prob = probs))
}) %>%
    bind_rows()

dist_of_dist %>%
    ggplot(aes(x, prob)) +
    geom_line(aes(group = factor(i)), size = 0.1, alpha = 0.2) +
    scale_y_continuous(limits = c(0, 3), 
                       expand = expansion(mult = c(0, 0.02))) +
    labs(x = "probability admit", 
         y = "density",
         title = "Posterior distribution of beta distributions")
```

- use posterior check to see how the beta distribution of probabilities of admissions influences predicted counts of applications admitted
    * the y-axis shows the predicted number of admits per row of the data frame on the x-axis
    * the purple dots are the actual values
    * open circles are the estimates from the model with 89% intervals
    * the `+` are the 89% interval of predicted counts of admission
        - shows there is a lot of dispersion
        - this is from the different departments, but the model doesn't know about them
        - the beta distribution accounts for this heterogeneity

```{r}
postcheck(m11_5)
```

### 11.3.2 Negative-binomial or gamma-Poisson

- *negative-binomial* (or *gamma-Poisson*): assumes that each Poisson count observation has its own rate
    * assumes the shape of the gamma distribution to describe the Poisson rates
    * predictor variables adjust the shape of this distribution, not the expected value of each observation
    * use the gamma distribution because it makes the math easier
- fitting a gamma-Poisson uses the `dgampois()` function
    * distribution is defined by a mean $\mu$ and scale $\theta$
    * as $\theta$ increases, the gamma distribution becomes for dispersed around the mean
- below are some examples of gamma distributions
    * as $\theta$ approaches 0, the gamma approaches a Gaussian

```{r}
mu <- 3
thetas <- c(1, 2, 5, 10, 20)
x_vals <- seq(0, 10, length.out = 50)
gamma_dist_res <- tibble()
for (theta in thetas) {
    gamma_dist_res <- bind_rows(
        gamma_dist_res,
        tibble(theta, 
               mu,
               x = x_vals,
               d = dgamma2(x_vals, mu, theta))
    )
}

gamma_dist_res %>%
    mutate(params = paste0("mu: ", mu, ", theta: ", theta)) %>%
    ggplot(aes(x = x, y = d)) +
    geom_line(aes(group = params, color = params)) +
    scale_color_brewer(palette = "Dark2") +
    theme(
        legend.position = c(0.7, 0.7)
    ) +
    labs(x = "probability", y = "density", color = NULL,
         title = "Gamma distributions")
```

- for fitting:
    * a linear model can be attached to the definition of $\mu$ using the log link function
    * there are examples in `?dgampois` and practice problems at the end of the chapter

### 11.3.3 Over-dispersion, entropy, and information criteria

- should use DIC instead of WAIC
    * see text for an explination of why
