---
title: "Chapter 12. Multilevel Models"
output: github_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE, 
                      cache = TRUE)

library(glue)
library(broom)
library(patchwork)
library(MASS)
library(rethinking)
library(tidyverse)
library(conflicted)


conflict_prefer("filter", "dplyr")
conflict_prefer("select", "dplyr")
conflict_prefer("rename", "dplyr")

# To be able to use `map2stan()`:
conflict_prefer("collapse", "dplyr")
conflict_prefer("extract", "tidyr")
conflict_prefer("rstudent", "rethinking")
conflict_prefer("lag", "dplyr")
conflict_prefer("map", "purrr")
conflict_prefer("Position", "ggplot2")

# To be able to use 'MASS':
conflict_prefer("area", "patchwork")

theme_set(theme_minimal())
source_scripts()
set.seed(0)

# For 'rethinking'.
options(mc.cores = parallel::detectCores())
rstan_options(auto_write = TRUE)

purple <- "#8067bf"
blue <- "#5896d1"
light_grey <- "grey80"
grey <- "grey50"
dark_grey <- "grey25"
```

- multi-level models remember features of each cluster in the data as they learn about all of the clusters
    * depending on the variation across clusters, the model pools information across clusters
    * *the pooling improves estimates about each cluster*
- benefits of the multilevel approach:
    1. improved estimates for repeat sampling
    2. improved estimates for imbalance in sampling
    3. estimates of variation
    4. avoid averaging and retain variation
- multilevel regression should be the default approach
- this chapter starts with the foundations and the following two are more advanced types of multilevel models

## 12.1 Example: Multilivel tadpoles

- example: Reed frog tadpole mortality
    * `surv`: number or survivors
    * `count`: initial number

```{r}
data("reedfrogs")
d <- as_tibble(reedfrogs)
skimr::skim(d)
```

- there is a lot of variation in the data
    * some from experimental treatment, other sources do exist
    * each row is a fish tank that is the experimental environment
    * each tank is a cluster variable and there are repeated measures from each
    * each tank may have a different baseline level of survival, but don't want to treat them as completely unrelated
        - a dummy variable for each tank would be the wrong solution
- *varying intercepts model*: a multilevel model that estimates an intercept for each tank and the variation among tanks
    * for each cluster in the data, use a unique intercept parameter, adaptively learning the prior common to all of the intercepts
    * what is learned about each cluster informs all the other clusters
- model for predicting tadpole mortality in each tank (nothing new)

$$
s_i \sim \text{Binomial}(n_i, p_i) \\
\text{logit}(p_i) = \alpha_{\text{tank}[i]} \\
\alpha_{\text{tank}} \sim \text{Normal}(0, 5) \\
$$

```{r}
d$tank <- 1:nrow(d)

m12_1 <- map2stan(
    alist(
        surv ~ dbinom(density, p),
        logit(p) <- a_tank[tank],
        a_tank[tank] ~ dnorm(0, 5)
    ),
    data = d
)

plot(m12_1)
precis(m12_1, depth = 2)
```

- can get expected mortality for each tank by taking the logistic of the coefficients

```{r}
logistic(coef(m12_1)) %>%
    enframe() %>%
    mutate(name = str_remove_all(name, "a_tank\\[|\\]"),
           name = as.numeric(name)) %>%
    ggplot(aes(x = name, y = value)) +
    geom_col() +
    scale_x_continuous(expand = c(0, 0)) +
    scale_y_continuous(expand = expansion(mult = c(0, 0.02))) +
    labs(x = "tank",
         y = "estimated probability survival",
         title = "Single-level categorical model estimates of tadpole survival")
```

- fit a multilevel model by adding a prior for the `a_tank` parameters as a function of its own parameters
    * now the priors have prior distributions, creating two *levels* of priors

$$
s_i \sim \text{Binomial}(n_i, p_i) \\
\text{logit}(p_i) = \alpha_{\text{tank}[i]} \\
\alpha_{\text{tank}} \sim \text{Normal}(\alpha, \sigma) \\
\alpha \sim \text{Normal}(0, 1) \\
\sigma \sim \text{HalfCauchy}(0, 1)
$$

```{r}
m12_2 <- map2stan(
    alist(
        surv ~ dbinom(density, p),
        logit(p) <- a_tank[tank],
        a_tank[tank] ~ dnorm(a, sigma),
        a ~ dnorm(0, 1),
        sigma ~ dcauchy(0, 1)
    ),
    data = d,
    iter = 4000,
    chains = 4,
    cores = 1
)

plot(m12_2)
print(m12_2)
precis(m12_2, depth = 2)
```
- interpretation:
    * $\alpha$: one overall sample intercept
    * $\sigma$: variance among tanks
    * 48 per-tank intercepts

```{r}
compare(m12_1, m12_2)
```

- from the comparison, see that the multilevel model only has ~38 effective parameters
    * 12 fewer than the single-level model because the prior assigned to each intercept shrinks them all towards the mean $\alpha$
        - *$\alpha$ is acting like a regularizing prior, but it has been learnt from the data*
- plot and compare the posterior medians from both models

```{r}
post <- extract.samples(m12_2)

d %>%
    mutate(propsurv_estimate = logistic(apply(post$a_tank, 2, median)),
           pop_size = case_when(
               density == 10 ~ "small tank",
               density == 25 ~ "medium tank",
               density == 35 ~ "large tank"
           ),
           pop_size = fct_reorder(pop_size, density)) %>%
    ggplot(aes(tank)) +
    facet_wrap(~ pop_size, nrow = 1, scales = "free_x") +
    geom_hline(yintercept = logistic(median(post$a)), 
               lty = 2, color = dark_grey) +
    geom_linerange(aes(x = tank, ymin = propsurv, ymax = propsurv_estimate), 
                   color = light_grey, size = 1) +
    geom_point(aes(y = propsurv), 
               color = grey, size = 1) +
    geom_point(aes(y = propsurv_estimate), 
               color = purple, size = 1) +
    labs(x = "tank",
         y = "proportion surivival",
         title = "Propotion of survival among tadpoles from different tanks.")
```

c
