---
title: "Chapter 4. Linear Models"
output: github_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)

library(glue)
library(conflicted)
library(rethinking)
library(patchwork)
library(tidyverse)

theme_set(theme_minimal())

conflict_prefer("filter", "dplyr")

set.seed(0)
```

- correlation is very common in the real world
- *multivariate regresssion* is using more than one predictor variable to model an outcome
- reasons to use multivariate regression:
    * a way of "controlling" for confounding variables
    * multiple causation
    * interaction between variables
- this chapter focuses on two thing multivate models can help with:
    * revealing spurious correlations
    * revealing important correlations masked by hidden correlations of other variables
- this chapter will also discuss:
    * multicolinearity
    * categorical variables

## 5.1 Spurious association

- example: correlation between divorce rate and marriage rate
    * need to be married to get divorced
    * perhaps higher rates of marriage indicate that marriage is more important, leading to fewer divorces
    * another predictor: median age at marriage
- we can fit a model of median age predicting divorce rate
    * this is the same as in the previous chapter
    * $D_i$: divorce rate for state $i$; $A_i$: median age at marriage in state $i$

$$
D_i \sim \text{Normal}(\mu_i, \sigma) \\
\mu_i = \alpha + \beta_A A_i \\
\alpha \sim \text{Normal}(10, 10) \\
\beta_A \sim \text{Normal}(0, 1) \\
\sigma \sim \text{Uniform}(0, 10)
$$

```{r}
# Load data.
data("WaffleDivorce")
d <- WaffleDivorce

# Stadardize predictor.
d$MedianAgeMarriage_std <- (d$MedianAgeMarriage - mean(d$MedianAgeMarriage)) / sd(d$MedianAgeMarriage)

m5_1 <- quap(
    alist(
        Divorce ~ dnorm(mu, sigma),
        mu <- a + bA * MedianAgeMarriage_std,
        a ~ dnorm(10, 10),
        bA ~ dnorm(0, 1),
        sigma ~ dunif(0, 10)
    ),
    data = d
)

summary(m5_1)
```

```{r}
mam_seq <- seq(-3, 3.5, length.out = 30)

mu <- link(m5_1, data = data.frame(MedianAgeMarriage_std = mam_seq))
mu_map <- apply(mu, 2, chainmode)
m5_1_pred <- apply(mu, 2, PI) %>%
    t() %>%
    as.data.frame() %>%
    set_names(c("pi_5", "pi_94")) %>%
    as_tibble() %>%
    mutate(mam_std = mam_seq,
           mu_map = mu_map)

d %>%
    ggplot() +
    geom_point(aes(x = MedianAgeMarriage_std, y = Divorce)) +
    geom_ribbon(data = m5_1_pred,
                aes(x = mam_std, ymin = pi_5, ymax = pi_94),
                fill = "black", alpha = 0.2, color = NA) +
    geom_line(data = m5_1_pred,
                aes(x = mam_std, y = mu_map),
                color = "blue", alpha = 0.6, lty = 2, size = 1.3)
```

- and we can model the divorce rate on the number of mariages in a state:
    * $R_i$: rate of marriage in state $i$

$$
D_i \sim \text{Normal}(\mu_i, \sigma) \\
\mu_i = \alpha + \beta_R R_i \\
\alpha \sim \text{Normal}(10, 10) \\
\beta_A \sim \text{Normal}(0, 1) \\
\sigma \sim \text{Uniform}(0, 10)
$$

```{r}
# Stadardize predictor.
d$Marriage_std <- (d$Marriage - mean(d$Marriage)) / sd(d$Marriage)

m5_2 <- quap(
    alist(
        Divorce ~ dnorm(mu, sigma),
        mu <- a + bR * Marriage_std,
        a ~ dnorm(10, 10),
        bR ~ dnorm(0, 1),
        sigma ~ dunif(0, 10)
    ),
    data = d
)

summary(m5_2)
```

- but individual single-variate models cannot tell us which variable is more important or if they cancel each other out
- the question we want to answer: *"What is the predictive value of a variable, once I already know all of the other predictor variables?"*
    * after I know the marriage rate, what additional value is there in also knowing the age at marraige?
    * after I know the age at marriage, what additional value is there in also knowing the marriage rate?

### 5.1.1 Multivariate notation

- the strategy for building a multivariate model:
    1. nominate the predictor variables you want in the linear model of the mean
    2. for each predictor, make a parameter that will measure its association with the outcome
    3. multiply the parameter by the variable and add that term to the linear model
- the formula for our multiate model example on divorce rate:

$$
D_i \sim \text{Normal}(\mu_i, \sigma) \\
\mu_i = \alpha + \beta_R R_i + \beta_A A_i \\
\alpha \sim \text{Normal}(10, 10) \\
\beta_A \sim \text{Normal}(0, 1) \\
\beta_R \sim \text{Normal}(0, 1) \\
\sigma \sim \text{Uniform}(0, 10)
$$

- what does $\mu_i = \alpha + \beta_R R_i + \beta_A A_i$ mean:
    * the expected outcome for any state with marriage rate $R_i$ and median age at marriage $A_i$ is the sum of three independent terms
    * $\alpha$ is a constant that every state gets
    * $\beta_R R_i$ is the marriage rate multiplied against a coefficient $\beta_R$ that measures the association between the marriage rate and divorce rate
    * $\beta_A A_i$ is simillar to the second term, but for the association with median age at marriage

### 5.1.2 Fitting the model

- we can use the quadratic approximation to fit the model

```{r}
m5_3 <- quap(
    alist(
        Divorce ~ dnorm(mu, sigma),
        mu <- a + bR * Marriage_std + bA * MedianAgeMarriage_std,
        a ~ dnorm(10, 10),
        bA ~ dnorm(0, 1),
        bR ~ dnorm(0, 1),
        sigma ~ dunif(0, 10)
    ),
    data = d
)

summary(m5_3)
```

```{r}
plot(summary(m5_3))
```

- now, the coefficient for the marriage rate predictor is about zero and the coefficient of the median age is confidently below zero
    we can interpret these to mean: *"Once we know the median age at marriage for a state, there is little predictive power in also knowing the rate of marriage in that state."*
- we can make some plots to investigate how the model came to this conclusion

### 5.1.3 Plotting the multivariate posteriors

- we will use three types of interpretive plots
    1. *predictor residual plots*: show the outcome against residual predictor values
    2. *counterfactual plots*: show the implied predictions for imaginary experiments in which the different predictor variables can be changed independently of one another
    3. *posterior prediction plots*: show model-based predictions against raw data, or otherwise diplay the error in prediction

#### 5.1.3.1 Predictor residual plots

- *predictor variable residual*: the average prediction error when using all other predictor variables to model a predictor of interest
    * plotting this against the outcome shows something like a bivariate regression that has already been "controlled" for all of the other predictors
    * it leaves the variation not expected by the model of the mean of the other predictors
- this is best illustrated by an example:
    * we will model the marriage rate using the meadian age at marriage

$$
R_i \sim \text{Normal}(\mu_i \sigma) \\
\mu_i = \alpha + \beta A_i \\
\alpha \sim \text{Normal}(0, 10) \\
\beta \sim \text{Normal}(1, 0) \\
\sigma \sim \text{Uniform}(0, 10)
$$

- since we are using centered variables, $\apha$ should be zero

```{r}
m5_4 <- quap(
    alist(
        Marriage_std ~ dnorm(mu, sigma),
        mu <- a + b*MedianAgeMarriage_std,
        a ~ dnorm(0, 10),
        b ~ dnorm(0, 1),
        sigma ~ dunif(0, 10)
    ),
    data = d
)

summary(m5_4)
```

- we then compute the residuals by subtracting the observed marriage rate in each state from the predicted rate when using age at marriage
    * a positive residual means the observed rate was in excess of that expected given the median age in that state

```{r}
mu <- coef(m5_4)["a"] + coef(m5_4)["b"] * d$MedianAgeMarriage_std
m_resid <- d$Marriage_std - mu
str(m_resid)
```

```{r}
d %>%
    mutate(mu = mu,
           resid = m_resid,
           resid_diff = mu + m_resid) %>%
    ggplot() +
    geom_linerange(aes(x = MedianAgeMarriage_std, 
                       ymin = mu, ymax = resid_diff),
                   size = 0.8, color = "grey60") +
    geom_point(aes(x = MedianAgeMarriage_std, y = Marriage_std),
               color = "black", size = 2) +
    geom_line(aes(x = MedianAgeMarriage_std, y = mu),
              color = "tomato", size = 1.3, alpha = 0.7) +
    labs(title = "Residual marriage rate estimated using the median age at marriage",
         subtitle = "The red line is the estimate, and the vertical lines are the residuals")
```

- we can then plot these residuals against the divorce rate
    * this is the linear relationship between divorce and marriage rates after "controlling" for median age of marriage
    
```{r}
d %>%
    mutate(mu = mu,
           resid = m_resid,
           resid_diff = mu + m_resid) %>%
    ggplot() +
    geom_point(aes(x = resid, y = Divorce),
               color = "black", size = 2) +
    geom_vline(xintercept = 0, lty = 2, color = "dodgerblue", size = 0.9) +
    labs(x = "residual marriage rate",
         title = "Residual marriage rate and Divorce",
         subtitle = "The linear relationship of marriage rate and divorce after correcting for median age at marriage")
```

- we can do the same caluclation in the other direction: find the residual of the median age modeled on the rate

```{r}
m5_4_2 <- quap(
    alist(
        MedianAgeMarriage_std ~ dnorm(mu, sigma),
        mu <- a + b*Marriage_std,
        a ~ dnorm(0, 10),
        b ~ dnorm(0, 1),
        sigma ~ dunif(0, 10)
    ),
    data = d
)

summary(m5_4)
```

```{r}
mu <- coef(m5_4_2)["a"] + coef(m5_4_2)["b"] * d$Marriage_std
m_resid <- d$MedianAgeMarriage_std - mu


p1 <- d %>%
    mutate(mu = mu,
           resid = m_resid,
           resid_diff = mu + m_resid) %>%
    ggplot(aes(x = Marriage_std)) +
    geom_linerange(aes(ymin = mu, ymax = resid_diff),
                   size = 0.8, color = "grey60") +
    geom_point(aes(y = MedianAgeMarriage_std),
               color = "black", size = 2) +
    geom_line(aes(y = mu),
              color = "tomato", size = 1.3, alpha = 0.7) +
    labs(title = "Residual median age estimated\nusing the marriage rate",
         subtitle = "The red line is the estimate,\nand the vertical lines are the residuals")

p2 <- d %>%
    mutate(mu = mu,
           resid = m_resid,
           resid_diff = mu + m_resid) %>%
    ggplot() +
    geom_point(aes(x = resid, y = Divorce),
               color = "black", size = 2) +
    geom_vline(xintercept = 0, lty = 2, color = "dodgerblue", size = 0.9) +
    labs(x = "residual median age",
         title = "Residual median age and Divorce",
         subtitle = "The linear relationship of median age\nand divorce after correcting for marriage rate")

p1 | p2
```

- the negative slope of the residual median age vs. Divorce (on the right in the above plot) indicates that the median age contains information even after adjusting for marriage rate

#### 5.1.3.2 Counterfactual plots

