---
title: "Chapter 8. Markov Chain Monte Carlo"
output: github_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)

library(glue)
library(broom)
library(patchwork)
library(rethinking)
library(tidyverse)
library(conflicted)


conflict_prefer("filter", "dplyr")
conflict_prefer("select", "dplyr")
conflict_prefer("rename", "dplyr")

# To be able to use `map2stan()`:
conflict_prefer("collapse", "dplyr")
conflict_prefer("extract", "tidyr")
conflict_prefer("rstudent", "rethinking")
conflict_prefer("lag", "dplyr")
conflict_prefer("map", "purrr")
conflict_prefer("Position", "ggplot2")

theme_set(theme_minimal())
source_scripts()
set.seed(0)
```

- estimation of posterior probability distributions using a stochastic process called *Markov chain Monte Carlo (MCMC)** estimation
    * sample directly from the posterior instead of approximating curves (like the quadratic approximation)
    * allows for models that do not assume a multivariate normality
        - can use generalized linear and multilevel models
- use **Stan* to fit these models

## 8.1 Good King Markov and His island kingdom

- a tale of the Good King Markov
    * he was kings of a ring of 10 islands
    * the second island had twice the population of the first, the third had three times the populaiton of the first, etc.
    * the king wanted to visit the islands in proportion to the population size, but didn't want to have to keep track of a schedule
    * he would also only travel between adjacent islands
    * he used the *Metropolis algorithm* to decide which island to visit next
        1. the king decides to stay or travel by a cion flip
        2. if the coin is heads, the king considers moving clockwise; if tails, he considers moving counter-clockwise; call this next island the "proposal" island
        3. to decide whether not to move, the king collects the number of seashells in proportion to the population size of the proposed island, and collects the number of stones relative to the population of the current island
        4. if there are more seashells, the king moves to the poposed island; else he discards the number of stones equal to the number of seashells and randomly selects from the remaining seashells and stones; if he selects a shell, he travels to the proposed island, else he stays
- below is a simulation of this process

```{r}
set.seed(0)

current_island <- 10
num_weeks <- 1e5
visited_islands <- rep(0, num_weeks)

flip_coin_to_decide_proposal_island <- function(x) {
    a <- sample(c(0, 1), 1)
    if (a == 0) {
        y <- x - 1
    } else {
        y <- x + 1
    }
    
    if (y == 0) {
        y <- 10
    } else if (y == 11) {
        y <- 1
    }
    return(y)
}

for (wk in seq(1, num_weeks)) {
    visited_islands[[wk]] <- current_island
    
    proposal_island <- flip_coin_to_decide_proposal_island(current_island)
    if (proposal_island > current_island) {
        current_island <- proposal_island
    } else {
        shells <- rep("shell", proposal_island)
        stones <- rep("stone", current_island - proposal_island)
        selection <- sample(unlist(c(shells, stones)), 1)
        if (selection == "shell") {
            current_island <- proposal_island
        }
    }
}
```

```{r}
tibble(wk = seq(1, num_weeks),
       islands = visited_islands) %>%
    ggplot(aes(x = factor(islands))) + 
    geom_bar(fill = "skyblue4") +
    labs(x = "islands",
         y = "number of weeks visited",
         title = "Distribution of King Markov's visits")
```

```{r}
tibble(wk = seq(1, num_weeks),
       islands = visited_islands) %>%
    slice(1:400) %>%
    ggplot(aes(x = wk, y = islands)) +
    geom_line(color = "skyblue4", alpha = 0.5) +
    geom_point(color = "skyblue4", size = 0.7) +
    scale_y_continuous(breaks = c(1:10)) +
    labs(x = "week number", y = "island",
         title = "The path of King Markov",
         subtitle = "Showing only the first 400 weeks")
```

- this algorithm still works if the king is equally likely to propose a move to any island from the current island
    * still use the proportion of the islands' populations as the probability of moving
- at any point, the king only needs to know the population of the current island and the proposal island

## 8.2 Markov chain Monte Carlo

- the *Metropolis algorithm* used above is an example of a *Markov chain Monte Carlo*
    * the goal is to draw samples from an unknown and complex target distribution
        - "islands": the parameter values (they can be continuous, too)
        - "population sizes": the posterior probabilities at each parameter value
        - "week": samples taken from the joint posterior of the parameters
    * we can use the samples from the Metropolis algorithm just like any other sampled distributions so far
- we will also cover *Gibbs sampling* and *Hamiltonian Monte Carlo*

### 8.2.1 Gibbs sampling

- can achieve a more efficient sampling procedure
    * *adaptive proposals*: the distribution of proposed parameter values adjusts itself intelligently depending upon the parameter values at the moment
    * the adaptive proposals depend on using particular combinations of prior distributions and likelihoods known as conjugate pairs
        - these pairs have analytical solutions for the posterior of an individual parameter
        - can use these solutions to make smart jumps around the joint posterior
- Gibbs sampling is used in BUGS (Bayesian inference Using Gibbs Sampling) and JAGS (Just Another Gibbs Sampler)
- limitations:
    * don't always want to use the conjugate priors
    * Gibbs sampling becomes very innefficient for models with hundreds or thousands of parameters

### 8.2.2 Hamiltonian Monte Carlo (HMC)

- more computationally expensive than Gibbs sampling and the Metropolis algorithm, but more efficient
    * doesn't require as many samples to describe the posterior distribution
- an analogy using King Monty:
    * king of a continuous stretch of land, not discrete islands
    * wants to visit his citizens in proportion to their local density
    * decides to travel back and forth across the length of the country, slowing down the vehicle when houses grow more dense, and speeing up when the houses are more sparse
    * requires knowledge of how quickly the population is changing at the current location
- in this analogy:
    * "royal vehicle": the current vector of parameter values
    * the log-posterior forms a bowl with the MAP at the nadir
    * the vehicle sweeps across the bowl, adjust the speed in proportion to the height on the bowl
- HMC does run a sort of physics simultation
    * a vector of parameters gives the position of a frictionless particle
    * the log-posterior provides the surface
- limitations:
    * HMC requires continuous parameters
    * HMC needs to be tuned to a particular model and its data
        - the particle needs mass (so it can have momentum)
        - other hyperparameters of MCMC is handeled by Stan

## 8.3 Easy HMC: `map2stan`

- `map2stan()` provides an interface to Stan similar to how we have used `quap()`
    * need to preprocess and variables transformations
    * input data frame can only have columns used in the formulae
- example: look at terrain ruggedness data

```{r}
data("rugged")
dd <- as_tibble(rugged) %>%
    mutate(log_gdp = log(rgdppc_2000)) %>%
    filter(!is.na(rgdppc_2000))
```

- fit a model to predict log-GDP with terrain ruggedness, continent, and the interaction of the two
- first fit with `quap()` like before

```{r}
m8_1 <- quap(
    alist(
        log_gdp ~ dnorm(mu, sigma),
        mu <- a + bR*rugged + bA*cont_africa + bAR*rugged*cont_africa,
        a ~ dnorm(0, 100),
        bR ~ dnorm(0, 10),
        bA ~ dnorm(0, 10),
        bAR ~ dnorm(0, 10),
        sigma ~ dunif(0, 10)
    ),
    data = dd
)
precis(m8_1)
```

### 8.3.1 Preparation

- must:
    * do any transformations beforehand (e.g. logarithm, squared, etc.)
    * reduce data frame to only used variables

```{r}
dd_trim <- dd %>%
    select(log_gdp, rugged, cont_africa)
```

### 8.3.2 Estimation

- can now fit the model with HMC using `map2stan()`

```{r}
m8_1stan <- map2stan(
    alist(
        log_gdp ~ dnorm(mu, sigma),
        mu <- a + bR*rugged + bA*cont_africa + bAR*rugged*cont_africa,
        a ~ dnorm(0, 100),
        bR ~ dnorm(0, 10),
        bA ~ dnorm(0, 10),
        bAR ~ dnorm(0, 10),
        sigma ~ dcauchy(0, 2)
    ),
    data = dd_trim
)

precis(m8_1stan)
```

- a half-Cauchy prior was used for $\sigma$
    * a uniform distribution would work here, too
    * it is a useful, "thick-tailed" probability
    * related to the Student $t$ distribution
    * can think of it as a weakly-regularizing prior for the standard deviation
- the estimates from this model are simillar to those from the quadratic prior
- a few differences to note about the output of `precis()` (`summary()`):
    * the *highest probability density intervals* (HPDI) are shown, not just percentiles intervals (PI), like before
    * two new columns (they will be discussed further later):
        - `n_eff`: a crude estimate of the number of independent samples that were collected
        - `Rhat4`: an estimate of the convergence of the Markov chains (1 is good)

### 8.3.3 Sampling again, in parallel
