---
title: "'tidybayes'"
output: github_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE, comment = "#>", dpi = 500)

library(glue)
library(mustashe)
library(magrittr)
library(broom)
library(patchwork)
library(rstanarm)
library(tidybayes)
library(tidyverse)
library(conflicted)


conflict_prefer("filter", "dplyr")
conflict_prefer("select", "dplyr")
conflict_prefer("rename", "dplyr")

# To be able to use `map2stan()`:
conflict_prefer("collapse", "dplyr")
conflict_prefer("extract", "tidyr")
conflict_prefer("lag", "dplyr")
conflict_prefer("map", "purrr")
conflict_prefer("Position", "ggplot2")

# To be able to use 'MASS':
conflict_prefer("area", "patchwork")

theme_set(theme_minimal())

# Turn a PI into a tibble.
pi_to_df <- function(list_pi) {
    list_pi %>%
        t() %>%
        as.data.frame() %>%
        janitor::clean_names() %>%
        as_tibble()
}


set.seed(0)


purple <- "#8067bf"
blue <- "#5896d1"
red <- "#d15858"
light_grey <- "grey80"
grey <- "grey50"
dark_grey <- "grey25"

# To shut-up `summarise()` in dplyr 1.0.0
options(dplyr.summarise.inform = FALSE)
```

['tidybayes' documentation website.](https://mc-stan.org/rstanarm/index.html)  
[CRAN : 'tidybayes'](https://cran.r-project.org/package=tidybayes)  
[A link to all 'tidybayes' vignettes.](https://mjskay.github.io/tidybayes/articles/)

---

## [Extracting and visualizing tidy draws from rstanarm models](https://mjskay.github.io/tidybayes/articles/tidy-rstanarm.html)

- example data set of 10 observations from 5 conditions each

```{r}
set.seed(5)
n <- 10
n_condition <- 5
ABC <- tibble(
    condition = rep(c("A","B","C","D","E"), n),
    response = rnorm(n * 5, c(0,1,2,1,-1), 0.5)
)

head(ABC)
```

```{r}
ABC %>%
    ggplot(aes(y = condition, x = response)) +
    geom_point()
```

### Model

- hierarchical model with shrinkage towards a global mean

```{r}
stash("m", depends_on = "ABC", {
    m <- stan_lmer(
        response ~ (1|condition), data = ABC,
        prior = normal(0, 1, autoscale = FALSE),
        prior_aux = student_t(3, 0, 1, autoscale = FALSE),
        adapt_delta = .99,
        refresh = 0
    )
})

m
```

### Extracting draws from a fit in tidy-format using `spread_draws()`

- `get_variables()` function to list the raw model variable names
    - `(Intercept)` is the global mean and the `b` parameters are offsets from that mean for each condition

```{r}
get_variables(m)
```

- `spread_draws()` function to extract posterior samples for the variables
    - includes a specification format that can be used to extract model variables and their indices

```{r}
m %>%
    spread_draws(b[term,group]) %>%
    head()
```

```{r}
# Can remove the `term` variable because there is only an `(Intercept)` variable
# in the model.
m %>%
    spread_draws(b[,group]) %>%
    head()
```

```{r}
m %>%
    spread_draws(b[,group]) %>% 
    separate(group, c("group", "condition"), sep = ":") %>%
    head()
```

- use the `sep` argument to parse the variable names further

```{r}
m %>%
    spread_draws(b[,group,condition], sep = "[, :]") %>%
    head()
```

### Point summaries and intervals

- family of functions for point summaries of posterior distributions
    - follow the naming scheme `[median|mean|mode]_[qi|hdi]()`

```{r}
m %>%
    spread_draws(`(Intercept)`, sigma) %>%
    median_qi(`(Intercept)`, sigma)
```

- can use `gather_draws()` to get a long format

```{r}
m %>%
    gather_draws(`(Intercept)`, sigma) %>%
    head()
```

```{r}
m %>%
    gather_draws(`(Intercept)`, sigma) %>%
    median_qi()
```

```{r}
m %>%
    spread_draws(b[,group]) %>%
    median_qi()
```

### Posterior fits

- use `add_fitted_draws()` to mimic `rstanarm::posterior_linpred()`

```{r}
ABC %>%
    modelr::data_grid(condition) %>%
    add_fitted_draws(m) %>%
    head()
```

```{r}
ABC %>%
    modelr::data_grid(condition) %>%
    add_fitted_draws(m) %>%
    ggplot(aes(x = .value, y = condition)) +
    ggdist::stat_pointinterval(.width = c(0.50, 0.89))
```

```{r}
ABC %>%
    modelr::data_grid(condition) %>%
    add_fitted_draws(m) %>%
    ggplot(aes(x = .value, y = condition)) +
    ggdist::stat_halfeye(aes(fill = stat(abs(x) < 0.8)), 
                         .width = c(0.50, 0.89)) +
    geom_vline(xintercept = c(-0.8, 0.8), size = 1, lty = 2, color = dark_grey) +
    scale_fill_manual(values = c(blue, grey))
```

### Posterior predictions

- use `add_predicted()` to mimic `rstanarm::posterior_predict()` 

```{r}
ABC %>%
      modelr::data_grid(condition) %>%
      add_predicted_draws(m) %>%
      ggplot(aes(x = .prediction, y = condition)) +
      ggdist::stat_halfeye()
```

### Fit/prediction curves

```{r}
m_mpg <- stan_glm(mpg ~ hp * cyl, data = mtcars, refresh = 0)
```

- the following two examples show the models with the *probability bands*
    - uses `add_fitted_draws()`

```{r}
mtcars %>%
    group_by(cyl) %>%
    modelr::data_grid(hp = modelr::seq_range(hp, n = 51)) %>%
    add_fitted_draws(m_mpg) %>%
    head()
```

```{r}
mtcars %>%
    group_by(cyl) %>%
    modelr::data_grid(hp = modelr::seq_range(hp, n = 51)) %>%
    add_fitted_draws(m_mpg) %>%
    ggplot(aes(x = hp, y = mpg, color = factor(cyl))) +
    stat_lineribbon(aes(y = .value)) +
    geom_point(data = mtcars) +
    scale_fill_brewer(palette = "Greys") +
    scale_color_brewer(palette = "Set2")
```

```{r}
mtcars %>%
    group_by(cyl) %>%
    modelr::data_grid(hp = modelr::seq_range(hp, n = 101)) %>%
    add_fitted_draws(m_mpg, n = 100) %>%
    ggplot(aes(x = hp, y = mpg, color = factor(cyl))) +
    geom_line(aes(y = .value, group = paste(cyl, .draw)), 
              alpha = .1) +
    geom_point(data = mtcars) +
    scale_color_brewer(palette = "Set2")
```

- the following two examples show the models with the *posterior predictions*
    - uses `add_predicted_draws()`

```{r}
mtcars %>%
    group_by(cyl) %>%
    modelr::data_grid(hp = modelr::seq_range(hp, n = 101)) %>%
    add_predicted_draws(m_mpg) %>%
    ggplot(aes(x = hp, y = mpg, 
               color = factor(cyl), fill = factor(cyl))) +
    stat_lineribbon(aes(y = .prediction), 
                    .width = c(.95, .80, .50), alpha = 1/4) +
    geom_point(data = mtcars) +
    scale_fill_brewer(palette = "Set2") +
    scale_color_brewer(palette = "Dark2")
```

```{r}
mtcars %>%
    group_by(cyl) %>%
    modelr::data_grid(hp = modelr::seq_range(hp, n = 101)) %>%
    add_predicted_draws(m_mpg) %>%
    ggplot(aes(x = hp, y = mpg)) +
    stat_lineribbon(aes(y = .prediction), 
                    .width = c(.99, .95, .8, .5),
                    color = RColorBrewer::brewer.pal(5, "Blues")[[5]]) +
    geom_point(data = mtcars) +
    scale_fill_brewer() +
    facet_grid(. ~ cyl, space = "free_x", scales = "free_x")
```
